---
title: "Modeling the Spatial Variation of the Explanatory Factors of Water Point Status using Geographically Weighted Logistic Regression"
subtitle: "In-class_Ex5"
editor: visual
author: Yang Tianyi
execute: 
  warning: false
  message: false
format: html
---

# Setting the Scene

To build an explanatory model to discover factor affecting water point status in Osun State, Nigeria

# Model Variables

-   Dependent variable: Water point status (i.e.functional/non-functional)

-   Independent variables:

    -   distance_to_primary_road

    -   distance_to_secondary_road

    -   distance_to_tertiary_road

    -   distance_to_city

    -   distance_to_town,

    -   water_point_population

    -   local_population_1km

    -   usage_capacity

    -   is_urban

    -   water_source_clean

# Getting Started

The code chunks below installs and launches specific R packages into R including: sf, tidyverse, funModeling, blorr, corrplot, ggpubr, sf, spdep, GWmode, tmap, skimr, and caret.

```{r}
pacman::p_load(sf, tidyverse, funModeling, blorr, corrplot, ggpubr, sf, spdep, GWmodel, tmap, skimr, report, caret, lifecycle, ggplot2, lattice)
```

# Data Input

In this in-class exercise, two data sets will be used

## Importing the Data

```{r}
Osun <- read_rds("data/Osun.rds")
Osun_wp_sf <- read_rds("data/Osun_wp_sf.rds")
```

```{r}
Osun_wp_sf %>%
  freq(input = 'status')
```

```{r}
tmap_mode("view")
tm_shape(Osun)+
  tm_polygons(alpha = 0.4) + 
tm_shape(Osun_wp_sf)+
  tm_dots(col = "status",
          alpha = 0.6)+
  tm_view(set.zoom.limits = c(9,12))
```

## Exploratory Data Analysis

### Summary Statistics with skim

```{r}
Osun_wp_sf %>%
  skim()
```

```{r}
Osun_wp_sf_clean <- Osun_wp_sf %>%
  filter_at(vars(status,
                 distance_to_primary_road,
                 distance_to_secondary_road,
                 distance_to_tertiary_road,
                 distance_to_city,
                 distance_to_town,
                 water_point_population,
                 local_population_1km,
                 usage_capacity,
                 is_urban,
                 water_source_clean),
           all_vars(!is.na(.))) %>%
  mutate(usage_capacity = as.factor(usage_capacity))
```

## Correlation Analysis

```{r}
Osun_wp <- Osun_wp_sf_clean %>%
  select(c(7, 35:39, 42:43, 46:47,57)) %>%
  st_set_geometry(NULL)
```

```{r}
cluster_vars.cor = cor(
  Osun_wp[,2:7])
corrplot.mixed(cluster_vars.cor,
               lower = "ellipse",
               upper = "number",
               tl.pos = "lt",
               diag = "l",
               tl.col = "black")
```

```{r}
model <- glm(status ~ distance_to_primary_road+
               distance_to_secondary_road+
               distance_to_tertiary_road+
               distance_to_city+
               distance_to_town+
               is_urban+
               usage_capacity+
               water_source_clean+
               water_point_population+
               local_population_1km,
             data = Osun_wp_sf_clean,
             family = binomial(link = 'logit'))
```

```{r}
blr_regress(model)
```

```{r}
report(model)
```

```{r}
blr_confusion_matrix(model, cutoff = 0.5)
```

```{r}
Osun_wp_sp <- Osun_wp_sf_clean %>%
  select(c(status,
           distance_to_primary_road,
           distance_to_secondary_road,
           distance_to_tertiary_road,
           distance_to_city,
           distance_to_town,
           water_point_population,
           local_population_1km,
           usage_capacity,
           is_urban,
           water_source_clean)) %>%
  as_Spatial()
Osun_wp_sp
```

```{r}
bw.fixed <- bw.ggwr(status ~ 
               distance_to_primary_road+
               distance_to_secondary_road+
               distance_to_tertiary_road+
               distance_to_city+
               distance_to_town+
               water_point_population+
               local_population_1km+
               is_urban+
               usage_capacity+
               water_source_clean,
             data = Osun_wp_sp,
             family = "binomial",
             approach = "AIC",
             kernel = "gaussian",
             adaptive = FALSE,
             longlat = FALSE)
```

```{r}
bw.fixed
```

```{r}
gwlr.fixed <- ggwr.basic(status ~ 
               distance_to_primary_road+
               distance_to_secondary_road+
               distance_to_tertiary_road+
               distance_to_city+
               distance_to_town+
               water_point_population+
               local_population_1km+
               is_urban+
               usage_capacity+
               water_source_clean,
            data = Osun_wp_sp,
            bw = 2597.255,
            family = "binomial",
            kernel = "gaussian",
            adaptive = FALSE,
            longlat = FALSE)
```

```{r}
gwr.fixed <- as.data.frame(gwlr.fixed$SDF)
```

```{r}
gwr.fixed <- gwr.fixed %>%
  mutate(most = ifelse(
    gwr.fixed$yhat >= 0.5, T, F))
```

```{r}
library(lifecycle)
library(caret)
gwr.fixed$y <- as.factor(gwr.fixed$y)
gwr.fixed$most <- as.factor(gwr.fixed$most)
CM <- confusionMatrix(data=gwr.fixed$most, reference = gwr.fixed$y)
CM
```

### Visualizing gwLR

```{r}
Osun_wp_sf_selected <- Osun_wp_sf_clean %>%
  select(c(ADM2_EN, ADM2_PCODE,
           ADM1_EN, ADM1_PCODE,
           status))
```

```{r}
gwr_sf.fixed <- cbind(Osun_wp_sf_selected, gwr.fixed)
```

```{r}
tmap_mode("view")
prob_T <- tm_shape(Osun) + 
  tm_polygons(alpha = 0.1) + 
tm_shape(gwr_sf.fixed) + 
  tm_dots(col = "yhat",
          border.col = "gray60",
          border.lwd = 1) + 
  tm_view(set.zoom.limits = c(8,14))
prob_T
```

```{r}
#| eval = false
tertiary_TV <- tm_shape(Osun) + 
  tm_polygons(alpha = 0.1) +
tm_shape(gwr_sf.fixed) +
  tm_dots(col = "distance_to_tertiary_road_TX", 
          border.col = "gray60", 
          border.lwd = 1) +
  tm_view(set.zoom.limits = (8,14))
  
tmap_arrange(tertiary_SE, tertiary_TV, asp = 1, ncol = 2, sync = TRUE)
```
